Characterizing subarctic biomes for land surface modeling of pollution and climate risk

### 2021-04-16 VegOzone.tex
Add reference regarding CAMSRA
Comments and suggestions by Lisa Section 1.

ground-level ozone is a highly toxic pollutant known to cause damage to a variety of ecosystems around the World \citep{PT:Emberson2020}. At continuously high concentration levels, ozone negatively affects the life expectancy of humans \citep{WHO2008} and is a cause of visible injury, photosynthetic damage, early senescence as well as programmed cell death of plants \citep{PCE:Kangasjarvi2005}. Annual global yield losses of four major crops (wheat, rice, maize, and soybean) of $3-15\,\unit{\%}$ \citep{PJ:Ainsworth2017} and a suggested loss in primary production in forestry of $7\,\unit{\%}$ \citep{GCB:Wittig2009,EP:Matyssek2012} has been attributed to ozone.
The crop yield loss studies indicate a threat to food security in rapidly developing countries, e.g., in East and South-East Asia \citep{GCB:Tang2013,NCC:Tai2014,AE:Chuwah2015,GCB:Mills2018}, where ozone concentrations related to enhanced pollutant emissions from the transport, residential, power generation, and industrial sectors have increased in the past few decades. Even though long-term (pre-1950) observations of ground-level ozone concentrations are scarce; evidence from trend analysis of those few sites suggests that since the industrial revolution, background concentrations in the northern hemisphere have at least doubled and continued to increase until relatively recently \citep[Chapter 2]{IPCC2013}. Recent trend analyses indicate that a maximum was reached in 2007 \citep{AE:Derwent2018}, after which tropospheric ozone started to level off or even decline in Europe \citep{ESA:Cooper2014, ACP:Wespes2018,ESA:Gaudel2018}. This can be attributed to a successful implementation of air quality regulations coupled with economic restructuring which saw a move from coal to gas, reducing the number of episodes of peak concentrations especially in summertime over Europe and North America \citep[e.g.,][]{ESA:Fleming2018, ESA:Mills2018}. However, there are indications that an increase in the frequency of heatwave induced thermal stress will reduce the vegetations' capacity for ozone uptake and therefore diminish these achievements \citep{NCC:Lin2020}.\\

Ground-level ozone is formed in complex photochemical cycles involving precursor gases such as carbon monoxide (\chem{CO}) and hydrocarbons are known as volatile organic substances (VOCs) in the presence of nitrogen oxides (\chem{NO_x}). Despite a relatively short tropospheric lifetime of approximately $22\,\unit{days}$, ranging from a few days in the tropical boundary layer to up to $1\,\unit{year}$ in the upper troposphere \citep{JGR:Stevenson2005,ACP:Young2013}, ozone is subject to advection over long distances. Episodes of high ozone concentrations in the absence of local pollution sources can therefore often be attributed to the long-range transport of both ozone and its precursors. In northern Fennoscandia, episodes of enhanced ozone concentrations in 2003 and 2006 have been traced back to ozone precursor emissions related to forest fires in southern and eastern Europe, respectively \citep{AE:Lindskog2007,EP:Karlsson2013}. Besides, hydrocarbons are also emitted by vegetation itself in form of terpenes and monoterpenes, so-called biogenic volatile substances (BVOCs). These emissions are associated with, e.g., thermal stress, defense against herbivores, or even attraction of pollinators \citep{TPS:Penuelas2003}.

In the annual cycle in Boreal and also temperate climates, a distinct maximum of ground-level ozone concentrations typically occurs in spring followed by lower concentrations throughout the summer. Several conditions unique to the Arctic are stimulating the formation of this spring peak. Major pathways of ozone and its precursors' removal from the boundary layer include dry deposition on vegetated surfaces and photochemistry \citep{RG:Clifton2020}. As dry deposition to bare ground or snow and ice-covered surfaces was found to be low \citep{ACP:Helmig2007}, both pathways of removal are substantially suppressed during polar night \citep{AE:Monks2000}. This leads to a build-up of ozone and its precursors in winter. Come spring precursors are photochemically activated again, while snow cover typically continues to prevail and suppress ozone dry deposition.
On the other hand, very high ozone concentrations occurring over shorter periods can often be attributed to deep tropopause folding events over the Arctic during winter. These events facilitate an intrusion of stratospheric air masses which are enriched in ozone \citep{JGR:Skerlak2015}.
The spring peak is more pronounced in northern Fennoscandia than in more southern locations but shows a  high interannual variability \citep{AB:Klingberg2009, BER:Klingberg2019}. A cause for this is the complex interplay between atmospheric dynamics and chemistry \citep{AE:Laurila1996,BER:Hatakka2003} as described above as well as anthropogenic and biogenic activity \citep{AE:Rummukainen1996,AE:Simpson2002,QJRMS:Galbally2007,NGS:Schnell2009}. \citet{ACP:Andersson2017} showed in a modeling study focused on Fennoscandia that the variability of ozone concentrations in winter can be attributed mainly to changes in atmospheric background (transport) of ozone, while summertime abundance is mostly affected by emissions of precursors in the rest of Europe. Meteorology plays an important role throughout the year.

@@

Ozone acts as oxidative stress to plants. Its main action is imposed through reactions occurring in the cell walls and cell membranes of mesophyll cells inside the leaves. Ozone enters the leaves through stomata, which are kept open to allow for gas exchange allowing the entry of \chem{CO_2} for photosynthesis and loss of \chem{H_2O} vapor via the plant transpiration stream. Stomatal aperture, and hence stomatal conductance to these gases (including ozone) will vary over the day and growing season primarily to balance \chem{CO_2} uptake against \chem{H_2O} vapor loss. The higher the stomatal conductance, the higher the potential for ozone uptake. Stomatal conductance has been empirically linked to environmental factors such as air temperature $T$, photosynthetic photon flux density (PPFD), vapor pressure deficit (VPD), and soil water potential (SWP) as well as photosynthesis itself \cite[e.g.]{PTRS:Jarvis1976, BallBerry1987, Emberson2000, ICP:MappingManual2017}. Results from open-top chamber (OTC) experiments on downy birch (\emph{Betula pubescens}) and mountain birch (\emph{Betula pubescens toruosa}), native to subarctic regions, indicated reductions in both biomass, in root:shoot ratio, and visible leaf damage under elevated ozone treatment ($\left<\chem{[O_3]}\right> = 36-54\,\unit{ppb}$) \citep{Amb:Manninen2009}. Although considered to be more ozone tolerant due to an absence of visible injuries \citep{Amb:Girgzdiene2009}, \citet{Amb:Manninen2009} also found chlorophyll:carotenoid ratio and polyamines reductions in Scots pine (\emph{Pinus sylvestris}), indicating susceptibility to ozone in these species.\\

To measure the risk of damage inflicted to vegetation through ozone, different metrics have been defined. Two main types can be differentiated, exposure-based and uptake-based metrics, e.g., Accumulated Ozone Exposure over a threshold of $\mathrm{x}$ ($\mathrm{AOT_x}$) and Phyto-toxic Ozone Dose over a threshold $\mathrm{y}$ ($\mathrm{POD_y}$). Due to their simplicity, exposure-based metrics are still widely used. Comparisons between the two methods (e.g. $\mathrm{AOT_{40}}$ with $\mathrm{POD_3}$) show that uptake-based metrics are better suited to represent the actual risk to vegetation from ozone, especially in the climatically extreme parts of Europe \citep{GCB:Mills2010,ICP:MappingManual2017,iF:Juran2018} (e.g., Mediterranean, Baltics/Fennoscandia).

The exposure-based metric $\mathrm{AOT_x}$ (in units of $\unit{ppb}$) is defined as:
\begin{equation}
  \mathrm{AOT_x} = \int{\mathrm{max(\chem{[O_3]}-x)}\cdot \mathrm{dt}},
  \label{eq:aot}
\end{equation}
with the ambient ozone concentration $\chem{[O_3]}$. The threshold exeedence is typically integrated from $8\,\mathrm{am}-8\,\mathrm{pm}$ daily and $3\,\unit{months}$ (May--July) annually \citep{EUdirective2002}.

For computing the $\mathrm{POD_y}$, an estimate of the stomatal \chem{O_3} flux ($\Phi_\mathrm{sto}$) is calculated based on the assumption that the concentration of \chem{O_3} at the top of the canopy represents a reasonable estimate of the concentration at the upper surface of the laminar layer for a sunlit upper canopy leaf and environmental factors affecting stomatal conductance as described further in Section~\ref{sec:do3se}. $\mathrm{POD_y}$ is then calculated according to:
%
\begin{equation}
  \mathrm{POD_y} = \int{(\Phi_\mathrm{sto}-y)\cdot \mathrm{dt}},
  \label{eq:pod}
\end{equation}
%
with the hourly averaged ground-level ozone flux through the stomata Eq.~(\ref{eq:flux_stomata}) and a flux threshold $\mathrm{y}$ both given in units of \unit{nmol\,\chem{O_3}\,m^{-2}\,\mathrm{PLA}\,s^{-1}}. The flux threshold $\mathrm{y}$ is representing the detoxification potential of the plant and is typically only exceeded during daylight hours (i.e. when global radiation is above $50\,\unit{W\,m^{-2}}$).

Finally, exceedance of the flux-based critical level (CL) is calculated by:
%
\begin{equation}
  \mathrm{CL_{exeed}} = \mathrm{POD_y} - \mathrm{CL}.
\end{equation}
%

Within the scope of this study to assess probable effects of conditions unique to northern Fennoscandia on ozone uptake by plants and the resultant enhanced susceptibility to ozone inflicted damage, we conducted ground-level ozone concentration measurements at the Norwegian Institute of Bioeconomy Research (NIBIO) Environment Centre Svanhovd in the growing seasons of 2018/19 in collaboration with the Norwegian Institute for Air Research (NILU). 
In 2018, we observed ozone damage on semi-natural vegetation (clover) and crops (tobacco) in the ozone garden at Svanhovd. In contrast, no such damage was found on the clovers in 2019, although it was found on the sensitive tobacco cultivar. This indicates that the vegetation may have been more affected by ozone in 2018 than in 2019. In terms of $\mathrm{AOT_{40}}$, the threshold for protection of vegetation of $3000\,\unit{ppb\,h}$ (integrated over the growing season) given by the EU~directive \citep{EUdirective2002} has been well exceeded at many ozone monitoring sites in Southern Norway in the 2018 growing season \citep{NILU2019} but not at Svanhovd.

According to a report by the Norwegian Meteorological institute \citep{MetNOR2019}, the summer of 2018 was the warmest and driest ever recorded in eastern, western, and southern Norway. In the north (including Finnmark), it was amongst the warmest on record -- favorable conditions for planetary boundary layer ozone formation.
An unusually weak and northward shifted jet stream allowed for a persistent high-pressure system above northern Europe, including Fennoscandia which blocked the pressure systems for several consecutive months. In the period May -- July, southern Norway had temperatures $4\,\unit{^\circ C}$ above normal. While southern Norway had only about $60\,\unit{\%}$ of normal precipitation, northern Norway as a whole had close to normal precipitation, but with local variations.
Thermal stress on vegetation was exceptional not only in large parts of Fennoscandia but also in much of Europe, where the influence of the high-pressure system extended even over five months (April/May, July--September). These conditions gave rise to massive forest fires in different parts of Europe and thus an increase in ozone precursors. Boreal wildfires emit in addition to \chem{CO_2} also \chem{CO} ($\chem{[CO]}/\chem{[CO_2]} \propto 6-13\,\unit{\%}$) and \chem{VOCs} ($\chem{[VOCs]}/\chem{[CO_2]} \propto 0.5-1.5\,\unit{\%}$) \citep{AE:Cofer1990}.
A total of 2079 forest fires of different sizes were registered in Norway in 2018, twice as many as in the preceding years 2016/17 \citep{DSB2019}, last accessed April 2020). In Sweden, about 500 fires had been reported (five times more than in a usual summer), and an estimated total of $25000\,\unit{hectare}$ burned down in the course of what became known as the most extensive fires in central Sweden (G\"{a}vleborgs, J\"{a}mtlands, and Dalarnas l\"{a}n) in younger history \citep{SOU2019}. Coincident peak \chem{[O_3]} is found in ozone monitoring data from Svanhovd but also over whole northern Fennoscandia in July (see Fig.~\ref{fig:data_svanvik_2018_2019} or Fig.~\ref{fig:ozone_reconstruction_2018_07}).
In the 2003 drought period, elevated ozone in Europe was promoted by a combination of various factors such as reduced cloud (increased solar radiation), reduced dry deposition and turbulent mixing due to the stagnant weather conditions, and increased BVOC emissions \citep{JGR:Solberg2018}. It is likely that these factors also contributed to enhanced \chem{[O_3]} in 2018.

The methods described above rely on accurate representation of plant physiological response to climate variables (i.e. the species or plant functional type-specific responses used to parameterize risk assessment models) and are most often parameterized using data for less extreme climates, i.e., temperate or continental climates; and to a lesser extent, Boreal and Mediterranean climates. Parameterization of these models for more extreme bio-geographical regions such as subarctic climates has not been performed and relies on assuming that less extreme Boreal parameterization will be applicable. In this study, conditions at Svanhovd for the growing seasons 2018/19 shall serve as a reference.
It is important to understand the implications of using these more generic parameterizations for more extreme climate regions. Here we explore how risk estimates might be over or underestimated using generic vs bespoke subarctic parameterization. we also explore the role \chem{[O_3]} variability and changing climate variables to understand how this risk might alter in the future with the advent of climate change. This allows both an understanding of how the threat from ozone may change over time but also which are the key aspects of plant physiology that might determine risk (and hence which parameters might require special attention to get right) both for pollution impact assessment, but also more generally for describing gas exchange response to changing climates that will influence biochemical cycling. This might be one key to solve long-standing issues of coupled earth system models (ESM) which often lack PFTs specifically representing subarctic conditions in their land surface model \citep{GMD:Poulter2015,JAMES:Lawrence2019}.\\

The observation site at NIBIO Environment Centre Svanhovd and all related data are presented in Section~\ref{sec:data}. In Section~\ref{sec:stats}, we derive climatologies for ground-level ozone concentration at Svanhovd and northern Fennoscandia based on in situ observations and also compute climatologies of near-surface temperature, precipitation, and global radiation for Svanhovd. Based on these, we evaluate anomalies of environmental key variables in 2018/19 by statistical means. The derived ground-level ozone concentration climatology for northern Fennoscandia will also be compared with available ground-level ozone reanalysis products. We use the $\mathrm{DO_3SE}$ model to estimate the ozone uptake by natural and semi-natural vegetation to investigate main environmental drivers for the observed ozone damage in 2018 (Section~\ref{sec:do3se}). In this context, we introduce the$\mathrm{DO_3SE}$ model equations and derive bespoke parameterizations for subarctic species to assess systematic uncertainties arising from the use of default parameterizations. In Section~\ref{sec:conc} finally, we will summarize our results, conclude, and give an outlook on further work.

@@

There are many probable explanations for the discovered differences in the climatologies derived from the three reanalyses. The model enhancements from MACC reanalysis to CAMSRA have been reported and discussed by \citet{ACP:Inness2019} on a global level. Amongst others, assimilation of ozone profiles from satellite retrieval and an upgraded ozone chemistry scheme has lead to the enhanced performance of the latter. The higher spatial resolution of the regional air quality models used in the CAMSRAQ ensemble compared to the global reanalyses captures small scale depletion and peak episodes in \chem{[O_3]} better and enhances the capability of the models concerning daily and seasonal cycles in ozone. The latter is, of course, also affected by upgrades in the OpenIFS used as meteorological forcing. However, we can assume that the main driver for the strikingly different results is the assimilated observational ozone data. Satellite retrievals resolve \chem{[O_3]} at the surface rather poorly and hence do not constrain the global models well enough \citep{ACP:Andersson2017}.

@@

\begin{table}[t]
  \caption{Bespoke temperature parameterizations. MM refers to mapping manual \citep{GCB:Mills2010,ICP:MappingManual2017}.}
  \label{tab:sensitivity_tests_temp}
  \begin{tabular}{llrrr}
    \tophline
    Species & type & $T_\mathrm{min}$ & $T_\mathrm{opt}$ & $T_\mathrm{max}$ \\
    \middlehline
    \multirow{3}{*}{Downy birch} & MM & 5 & 20 & \multirow{3}{*}{100}\\
    & Boreal & 0 & 10\\
    & Cold & 5 & 15\\
    \middlehline
    \multirow{3}{*}{Norway spruce} & MM & 0 & 20 & \multirow{3}{*}{100}\\
    & Boreal & 0 & 10 & \\
    & Cold & 0 & 15 & \\
    \middlehline
    \multirow{3}{*}{Perennial grassland} & MM & 10 & 24 & 36\\
    & Boreal & 0 & 10 & 24 \\
    & Cold & 5 & 15 & 36\\
    \bottomhline
    \end{tabular}
\end{table}

\begin{table}[t]
  \caption{Bespoke light parameterizations. MM refers to mapping manual \citep{GCB:Mills2010,ICP:MappingManual2017}.}
  \label{tab:sensitivity_tests_light}
  \begin{tabular}{llr}
    \tophline
    Species & type & $\alpha$\\
    \middlehline
    \multirow{3}{*}{Downy birch} & MM & 0.004 \\
    & PPFD0.8 & 0.005\\
    & PPFD1.2 & 0.004\\
    \middlehline
    \multirow{3}{*}{Norway spruce} & MM & 0.006 \\
    & PPFD0.8 & 0.008  \\
    & PPFD1.2 & 0.005  \\
    \middlehline
    \multirow{3}{*}{Perennial grassland} & MM & 0.011\\
    & PPFD0.8 & 0.014 \\
    & PPFD1.2 & 0.009 \\
    \bottomhline
    \end{tabular}
\end{table}

>>>>

ground-level ozone is a highly toxic pollutant known to cause damage to of humans health \citep{WHO2008} and to a variety of ecosystems around the World \citep{PT:Emberson2020}. Ozone has been found to be a cause of visible injury, photosynthetic damage, early senescence as well as programmed cell death of plants \citep{PCE:Kangasjarvi2005}. Annual global yield losses of four major crops (wheat, rice, maize, and soybean) of $3-15\,\unit{\%}$ \citep{PJ:Ainsworth2017} and a suggested loss in primary production in forestry of $7\,\unit{\%}$ \citep{GCB:Wittig2009,EP:Matyssek2012} has been attributed to ozone.
The crop yield loss studies indicate a threat to food security in rapidly developing countries, e.g., in East and South-East Asia \citep{GCB:Tang2013,NCC:Tai2014,AE:Chuwah2015,GCB:Mills2018}, where ozone concentrations related to enhanced pollutant emissions from the transport, residential, power generation, and industrial sectors have increased in the past few decades. Even though long-term (pre-1950) observations of ground-level ozone concentrations are scarce; evidence from trend analysis of those few sites suggests that since the industrial revolution, background concentrations in the northern hemisphere have at least doubled and continued to increase until relatively recently \citep[Chapter 2]{IPCC2013}. Recent trend analyses indicate that a maximum was reached in 2007 \citep{AE:Derwent2018}, after which tropospheric ozone started to level off or even decline in Europe \citep{ESA:Cooper2014, ACP:Wespes2018,ESA:Gaudel2018}. This can be attributed to a successful implementation of air quality regulations coupled with economic restructuring which saw a move from coal to gas, reducing the number of episodes of peak concentrations especially in summertime over Europe and North America \citep[e.g.,][]{ESA:Fleming2018, ESA:Mills2018}. However, changes in environmental conditions associated with climate change, such as an increase in the frequency of heatwaves, could negate the effectiveness of emission reductions in relation to ozone impacts and air quality\citep{NCC:Lin2020}.

Ground-level ozone is formed in complex photochemical cycles involving precursor gases such as carbon monoxide (\chem{CO}) and hydrocarbons are known as volatile organic compounds (VOCs) in the presence of nitrogen oxides (\chem{NO_x}). Despite a relatively short tropospheric lifetime of approximately $22\,\unit{days}$, ranging from a few days in the tropical boundary layer to up to $1\,\unit{year}$ in the upper troposphere \citep{JGR:Stevenson2005,ACP:Young2013}, ozone and is precursors are subject to advection over long distances. Episodes of high ozone concentrations in the absence of local pollution sources can therefore often be attributed to the long-range transport of both ozone and its precursors. In northern Fennoscandia, episodes of enhanced ozone concentrations in 2003 and 2006 have been traced back to ozone precursor emissions related to forest fires in southern and eastern Europe, respectively \citep{AE:Lindskog2007,EP:Karlsson2013}. Besides, hydrocarbons are also emitted by vegetation itself in the form of terpenes and monoterpenes, so-called biogenic volatile substances (BVOCs). These emissions are thought to be a response to other factors, e.g., thermal stress, defense against herbivores, or even attraction of pollinators \citep{TPS:Penuelas2003}.

In the annual cycle in Boreal and also temperate climates, a distinct maximum of ground-level ozone concentrations typically occurs in spring followed by lower concentrations throughout the summer. Several conditions unique to the Arctic are stimulating the formation of this spring peak. Major pathways of ozone and its precursors' removal from the boundary layer include dry deposition on vegetated surfaces and photochemistry \citep{RG:Clifton2020}. As dry deposition to bare ground or snow and ice-covered surfaces was found to be low \citep{ACP:Helmig2007}, both pathways of removal are substantially suppressed during polar night \citep{AE:Monks2000}. This leads to a build-up of ozone and its precursors during winter. Come spring precursors are photochemically activated again, while snow cover typically continues to prevail and suppress ozone dry deposition.
On the other hand, very high ozone concentrations occurring over shorter periods can often be attributed to deep tropopause folding events over the Arctic during winter. These events facilitate an intrusion of stratospheric air masses which are enriched in ozone \citep{JGR:Skerlak2015}.
The spring peak is more pronounced in northern Fennoscandia than in more southern locations but shows a  high interannual variability \citep{AB:Klingberg2009, BER:Klingberg2019}. A cause for this is the complex interplay between atmospheric dynamics and chemistry \citep{AE:Laurila1996,BER:Hatakka2003} as described above as well as anthropogenic and biogenic activity \citep{AE:Rummukainen1996,AE:Simpson2002,QJRMS:Galbally2007,NGS:Schnell2009}. \citet{ACP:Andersson2017} showed in a modeling study focused on Fennoscandia that the variability of ozone concentrations in winter can be attributed mainly to changes in atmospheric background (transport) of ozone, while summertime abundance is mostly affected by emissions of precursors in the rest of Europe. Meteorology plays an important role throughout the year.

@@

Ozone acts as oxidative stress to plants. Its main action is imposed through reactions occurring in the cell walls and cell membranes of mesophyll cells inside the leaves. Ozone enters the leaves through stomata, leaf pores that enable gas exchange, allowing the entry of \chem{CO_2} for photosynthesis and loss of \chem{H_2O} vapor via the plant transpiration stream. Stomatal aperture, and hence stomatal conductance to these gases (including ozone) will vary over the day and growing season primarily to balance \chem{CO_2} uptake against \chem{H_2O} vapor loss. The higher the stomatal conductance, the higher the potential for ozone uptake. Stomatal conductance has been empirically linked to environmental factors such as air temperature $T$, photosynthetic photon flux density (PPFD), vapor pressure deficit (VPD), and soil water potential (SWP) as well as photosynthesis itself \cite[e.g.]{PTRS:Jarvis1976, BallBerry1987, Emberson2000, ICP:MappingManual2017}. Results from open-top chamber (OTC) experiments on downy birch (\emph{Betula pubescens}) and mountain birch (\emph{Betula pubescens toruosa}), native to subarctic regions, indicated reductions in both biomass, in root:shoot ratio, and visible leaf damage under elevated ozone treatment ($\left<\chem{[O_3]}\right> = 36-54\,\unit{ppb}$) \citep{Amb:Manninen2009}. Although considered to be more ozone tolerant due to an absence of visible injuries \citep{Amb:Girgzdiene2009}, \citet{Amb:Manninen2009} also found chlorophyll:carotenoid ratio and polyamines reductions in Scots pine (\emph{Pinus sylvestris}), indicating susceptibility to ozone in these species.\\

A substantial body of evidence exists that suggests flux-based metrics, that relate stomatal ozone uptake to vegetation damage, are biologically more relevant for risk assessments than exposure-based metrics. This is because they are able to account particular species characteristics (i.e. physiology and phenology) as well as environmental conditions that can decouple the relationship between ozone concentration, ozone uptake, and consequent damage \citep{PT:Emberson2020}. These flux based metrics have been shown to be better suited to represent the actual risk to vegetation from ozone, especially in the climatically extreme parts of Europe \citep{EP:Simpson2007,GCB:Mills2010,ICP:MappingManual2017}. Most of these previous studie have focussed on the Mediterranean where soil moisture deficits are thought to be most influential in decoupling concentration from uptake. Relatively few studies have explored the situation in Northern European climates where low temperatures may be considered more likely to influence ozone uptake \citep[e.g.]{iF:Juran2018}.

To estimate the flux-based metric, also referred to as the Phytotoxic Ozone Dose over a threshold y ($\mathrm{POD_y}$), an estimate of the stomatal \chem{O_3} flux ($\Phi_\mathrm{sto}$) is calculated based on the assumption that the concentration of \chem{O_3} at the top of the canopy represents a reasonable estimate of the concentration at the upper surface of the laminar layer for a sunlit upper canopy leaf and environmental factors affecting stomatal conductance as described further in Section~\ref{sec:do3se}. $\mathrm{POD_y}$ is then calculated according to:
%
\begin{equation}
  \mathrm{POD_y} = \int{(\Phi_\mathrm{sto}-y)\cdot \mathrm{dt}},
  \label{eq:pod}
\end{equation}
%
with the hourly averaged stomatal ozone flux $\Phi_\mathrm{sto}$ (see Eq.~(\ref{eq:flux_stomata}) and a stomatal ozone flux threshold $\mathrm{y}$ both given in units of \unit{nmol\,\chem{O_3}\,m^{-2}\,\mathrm{PLA}\,s^{-1}}. The flux threshold $\mathrm{y}$ represents the detoxification potential of the plant and is typically only exceeded during daylight hours (i.e. when global radiation is above $50\,\unit{W\,m^{-2}}$). $\mathrm{POD_y}$ is given in \unit{mmol\,m^{-2}} over the growing season.

The $\mathrm{POD_y}$ metric is currently used in risk assessments under the United Nations Economic Commission for Europe (UNECE) Long Range Transboundary Air Pollution (LRTAP) Convention to identify those locations across Europe where vegetation is at risk from ozone. This is with a view to developing effects-based emission reduction policy that can target those ozone precursor emissions that are most influential in causing damage. To achieve this the concept of critical levels (CLs) has been developed and applied \citep{Maas2016}. Exceedance of the CL is used to identify those areas across Europe that would benefit from targeted Europe-wide emission reductions. Methods to estimate $\mathrm{POD_y}$ and CLs are defined by the UNECE LRTAP Convention and described in a Mapping Manual \citep{ICP:MappingManual2017}. The CL is calculated by: 
%
\begin{equation}
  \mathrm{CL_{exeed}} = \mathrm{POD_y} - \mathrm{CL}.
\end{equation}
%

An application of the Mapping Manual method was made by \citet{GCB:Mills2011} where exceedance of the CL was related to observations of ozone damage to a clover biomonitor. This study clearly demonstrated the improved ability of flux-based metrics (in comparison to exposure-based metrics) to identify the geographical distribution of the risk of vegetation to ozone damage and also showed that the ozone damage could extend into more northerly regions of Europe. A similar biomonitoring study was conducted at the Norwegian Institute of Bioeconomy Research (NIBIO) Environment Centre Svanhovd to assess whether ozone was capable of inducing damage to vegetation growing under northern Fennoscandia conditions. In 2018, we indeed observed ozone damage on clover and tobacco. In contrast, no such damage was found on the clovers in 2019.

According to a report by the Norwegian Meteorological institute \citep{MetNOR2019}, the summer of 2018 was the warmest and driest ever recorded in eastern, western, and southern Norway. In the north (including Finnmark), it was amongst the warmest on record -- favorable conditions for planetary boundary layer ozone formation.
An unusually weak and northward shifted jet stream allowed for a persistent high-pressure system above northern Europe, including Fennoscandia which blocked the pressure systems for several consecutive months. In the period May -- July, southern Norway had temperatures $4\,\unit{^\circ C}$ above normal. While southern Norway had only about $60\,\unit{\%}$ of normal precipitation, northern Norway as a whole had close to normal precipitation, but with local variations.
Thermal stress on vegetation was exceptional not only in large parts of Fennoscandia but also in much of Europe, where the influence of the high-pressure system extended even over five months (April/May, July--September). These conditions gave rise to massive forest fires in different parts of Europe and thus an increase in ozone precursors. Boreal wildfires emit in addition to \chem{CO_2} also \chem{CO} ($\chem{[CO]}/\chem{[CO_2]} \propto 6-13\,\unit{\%}$) and \chem{VOCs} ($\chem{[VOCs]}/\chem{[CO_2]} \propto 0.5-1.5\,\unit{\%}$) \citep{AE:Cofer1990}.
A total of 2079 forest fires of different sizes were registered in Norway in 2018, twice as many as in the preceding years 2016/17 \citep{DSB2019}, last accessed April 2020). In Sweden, about 500 fires had been reported (five times more than in a usual summer), and an estimated total of $25000\,\unit{hectare}$ burned down in the course of what became known as the most extensive fires in central Sweden (G\"{a}vleborgs, J\"{a}mtlands, and Dalarnas l\"{a}n) in younger history \citep{SOU2019}. Coincident peak \chem{[O_3]} is found in ozone monitoring data from Svanhovd but also over whole northern Fennoscandia in July (see Fig.~\ref{fig:data_svanvik_2018_2019} or Fig.~\ref{fig:ozone_reconstruction_2018_07}).
In the 2003 drought period, elevated ozone in Europe was promoted by a combination of various factors such as wildfires, reduced cloud cover (increased solar radiation), reduced dry deposition and turbulent mixing due to the stagnant weather conditions, and increased BVOC emissions \citep{JGR:Solberg2018}. It is likely that these factors also contributed to enhanced \chem{[O_3]} in 2018.

The methods described above rely on accurate representation of plant physiological response to climate variables (i.e. the species or plant functional type-specific responses used to parameterize risk assessment models) and are most often parameterized using data for less extreme climates, i.e., temperate or continental climates; and to a lesser extent, Boreal and Mediterranean climates. Parameterization of these models for more extreme bio-geographical regions such as subarctic climates has not been performed and relies on assuming that less extreme Boreal parameterization will be applicable. In this study, conditions at Svanhovd for the growing seasons 2018/19 shall serve as a reference for present and probable future conditions in northern Fennoscandia.
It is important to understand the implications of using these more generic parameterizations for more extreme climate regions. Here we explore how risk estimates might be over or underestimated using generic vs bespoke subarctic parameterization. We also explore the role of \chem{[O_3]} variability and changing climate variables to understand how this risk might alter in the future with the advent of climate change. This allows an understanding of how the threat from ozone may change over time and which are the key aspects of plant physiology that might determine risk. Hence, we determin which parameters require special attention to define accurately for pollution impact assessment and more generally for describing gas exchange response to changing climates that will influence biochemical cycling. This might be one key to solve long-standing issues of coupled earth system models (ESM) which often lack PFTs specifically representing subarctic conditions in their land surface models \citep{GMD:Poulter2015,JAMES:Lawrence2019}.\\

The observation site at NIBIO Environment Centre Svanhovd and all related data are presented in Section~\ref{sec:data}. In Section~\ref{sec:stats}, we derive climatologies for ground-level ozone concentration at Svanhovd and northern Fennoscandia based on in situ observations and also compute climatologies of near-surface temperature, precipitation, and global radiation for Svanhovd. Based on these, we evaluate anomalies of key environmental variables in 2018/19 by statistical means. The derived ground-level ozone concentration climatology for northern Fennoscandia will also be compared with available ground-level ozone reanalysis products. We use the $\mathrm{DO_3SE}$ model to estimate the ozone uptake by natural and semi-natural vegetation to investigate the main environmental drivers for the observed ozone damage in 2018 (Section~\ref{sec:do3se}). In this context, we introduce the $\mathrm{DO_3SE}$ model equations and derive bespoke parameterizations for subarctic species to assess systematic uncertainties arising from the use of default generic parameterizations established from less climatically extreme conditions. In Section~\ref{sec:conc} finally, we will summarize our results and discuss how this research can support improvements in future assessments of pollution risk and more generally bio-geochemical cycling.

@@

There are many probable explanations for the discovered differences in the climatologies derived from the three reanalyses. The model enhancements from MACC reanalysis to CAMSRA have been reported and discussed by \citet{ACP:Inness2019} on a global level. Amongst others, assimilation of ozone profiles from satellite retrieval and an upgraded ozone chemistry scheme has lead to the enhanced performance of the latter. The higher spatial resolution of the regional air quality models used in the CAMSRAQ ensemble compared to the global reanalyses captures small scale depletion and peak episodes in \chem{[O_3]} better and enhances the capability of the models concerning daily and seasonal cycles in ozone. The latter is, of course, also affected by upgrades in the OpenIFS used as meteorological forcing. However, we can assume that the main driver for the strikingly different results is the assimilated observational ozone data. Satellite retrievals resolve \chem{[O_3]} at the surface rather poorly and hence do not constrain the global models well enough \citep{ACP:Andersson2017}. \citet{ACPD:Barten2020} reported pronounced biases for CAMSRA in the high Arctic (e.g., Summit, Greenland) and attribute this to dry deposition to the ocean.

@@

\begin{table}[t]
  \caption{Bespoke temperature parameterizations. MM refers to mapping manual \citep{GCB:Mills2011,ICP:MappingManual2017}.}
  \label{tab:sensitivity_tests_temp}
  \begin{tabular}{llrrr}
    \tophline
    Species & type & $T_\mathrm{min}$ & $T_\mathrm{opt}$ & $T_\mathrm{max}$ \\
    \middlehline
    \multirow{3}{*}{Downy birch} & MM & 5 & 20 & \multirow{3}{*}{100}\\
    & Boreal & 0 & 10\\
    & Cold & 5 & 15\\
    \middlehline
    \multirow{3}{*}{Norway spruce} & MM & 0 & 20 & \multirow{3}{*}{100}\\
    & Boreal & 0 & 10 & \\
    & Cold & 0 & 15 & \\
    \middlehline
    \multirow{3}{*}{Perennial grassland} & MM & 10 & 24 & 36\\
    & Boreal & 0 & 10 & 24 \\
    & Cold & 5 & 15 & 36\\
    \bottomhline
    \end{tabular}
\end{table}

\begin{table}[t]
  \caption{Bespoke light parameterizations. MM refers to mapping manual \citep{GCB:Mills2011,ICP:MappingManual2017}.}
  \label{tab:sensitivity_tests_light}
  \begin{tabular}{llr}
    \tophline
    Species & type & $\alpha$\\
    \middlehline
    \multirow{3}{*}{Downy birch} & MM & 0.004 \\
    & PPFD0.8 & 0.005\\
    & PPFD1.2 & 0.004\\
    \middlehline
    \multirow{3}{*}{Norway spruce} & MM & 0.006 \\
    & PPFD0.8 & 0.008  \\
    & PPFD1.2 & 0.005  \\
    \middlehline
    \multirow{3}{*}{Perennial grassland} & MM & 0.011\\
    & PPFD0.8 & 0.014 \\
    & PPFD1.2 & 0.009 \\
    \bottomhline
    \end{tabular}
\end{table}
